---
title: "A new-user, active comparator design; Per-protocol analysis"
author: "Kevin Ying,Peirong Hao,Yizhe Xu"
date: "2025-07-21"
output: html_document
---

date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: reference.bib  
link-citations: yes
objective: To apply the target trial emulation framework to create a analysis-ready data set
vignette: >
  %\VignetteIndexEntry{Target Trial Emulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(2009)
options(warn = -1)
```

Per-protocol (PP) analysis is a method used in the statistical evaluation of clinical trials to assess the effect of an intervention under ideal conditions. Unlike the intention-to-treat (ITT) approach, which includes all randomized participants regardless of adherence, PP analysis includes only those participants who fully comply with the trial protocol. This typically means excluding individuals who did not receive the assigned intervention, who withdrew early, or who deviated significantly from study procedures.

The primary objective of per-protocol analysis is to estimate the efficacy of a treatment—how well it works when taken exactly as intended. While this can offer valuable insight into the potential benefits of an intervention, it may also introduce bias due to the non-random exclusion of participants. Therefore, PP analysis is often used as a complement to ITT analysis rather than a replacement, helping to provide a more nuanced understanding of treatment effects.

In regulatory and clinical decision-making contexts, per-protocol analysis can be particularly relevant when determining the maximum achievable benefit of a treatment under optimal adherence conditions. However, careful consideration must be given to its limitations, especially in terms of generalizability and internal validity.

# Load packages
```{r, message = FALSE, warning = FALSE}
library(TrialEmulation)
library(dplyr)
library(tidyr)
library(ggplot2)
library(survival)
library(survminer)
library(lubridate)
source("R/generate observed data.R")
```
Simulate data
```{r}
# Set seed
set.seed(412)
#simulate data
simdata0=generate.obs.data(n=1000,J=20,beta0.trt0=1,beta.adherence=2, beta0.outcome=-5, beta.trt.effect=-1.2, beta0.censoring=-1)
simdata=simdata0%>%select(id, time, X1, X2, X3, X4, age, A, Y, C) 
```
Now we start to create a analysis-ready from the simulated data. We will emulate a randomized clinic trial described below. 
Add a table here to describe the trial design. 

First, we create an eligible indicator. Our inclusion criteria are: 1)age >=50, 2)no history of Y by current time, and 3)no treatment in previous 2 years of current time
```{r}
simdata <- simdata %>%
  group_by(id) %>%
  mutate(eligible = as.integer(age >= 50 & cumsum(Y) == 0 & (lag(A)+lag(A,2)) == 0)) %>%
  ungroup()
```
Let's take a look on a sample. 
```{r}
simdata%>%filter(id==2)

```
We can see that:  
At time 0 through 6, his age is below 50 and so e is not eligible for the study.
At time 7 through 9, his age is over 50 and he has no history of Y but he has treatment within previous 2 years, so he is not eligible.
At time 10 through 13, his age is 50 and he has no history of Y and no treatment in previous 2 years, so he is eligible.
At time 14 and after, he has treatment in previous 2 years, so he is not eligible.
Now we identify the subjects who are eligible ever and the first time they are eligible
```{r}
eligible_info <- simdata %>%
  group_by(id) %>%
  summarise(first_eligible_time = min(time[eligible == 1], na.rm = TRUE),
            ever_eligible = any(eligible == 1)) %>%
  ungroup()
```
How many ever eligible? 
```{r}
table(eligible_info$ever_eligible)
```
Check a few of them. 
```{r}
eligible_info %>% filter(id %in% 1:3)
```
As we explained above, subject 2 is first eligible at time 10. And we can also see that subject 1 and 3 are never eligible. 

Now merge the information
```{r}
simdata <- simdata %>%
  left_join(eligible_info, by = "id") 
```

now we are ready to do the analysis
We will perform per protocol (PP) analysis  
add a link to help document of data_preparation()
```{r}
working_dir <- getwd()
prep_PP_data <- data_preparation(
 data = simdata,
 id = "id", period = "time", treatment = "A",
 outcome = "Y", eligible = "eligible",
 estimand_type = "PP",
 outcome_cov = ~ X1 + X2 + X3 + X4 + age,
 model_var = "assigned_treatment",
 switch_d_cov = ~ X1 + X2 + X3 + X4 + age
 + time_on_regime + I(time_on_regime^2),
 switch_n_cov = ~ X3 + X4 + time_on_regime + I(time_on_regime^2),
 use_censor_weights = TRUE, cense = "C",
 cense_d_cov = ~ X1 + X2 + X3 + X4 + age,
 cense_n_cov = ~ X3 + X4,
 pool_cense = "numerator",
 data_dir = working_dir, save_weight_models = TRUE,
 chunk_size = 500, separate_files = TRUE,
 quiet = TRUE)
adh.model.d0=prep_PP_data$switch_models$switch_d0 #more anotations here
adh.model.d1=prep_PP_data$switch_models$switch_d1
adh.model.n0=prep_PP_data$switch_models$switch_n0
adh.model.n1=prep_PP_data$switch_models$switch_n1
cens.model.d0=prep_PP_data$censor_models$cens_d0
cens.model.d1=prep_PP_data$censor_models$cens_d1
cens.model.n=prep_PP_data$censor_models$cens_pool_n
```


Check how these models are fitted

```{r}
model.pp.cens.n=readRDS("C:/Users/u0399057/Box/Private/Kevin Ying/TTE vignette/cense_model_pool_n.rds")
dt.pp.cens.n=model.pp.cens.n$data
check.cens.n=simdata%>%left_join(dt.pp.cens.n%>%mutate(flag=1,time=period)%>%select(id,time,flag))
sum(check.cens.n$flag, na.rm = T)==dim(dt.pp.cens.n)[1]
check.cens.n%>%filter(id==6)
simdata%>%filter(id==6)

model.pp.cens.d0=readRDS("/Users/kevinying/TTEvignette/cense_model_d0.rds")
dt.pp.cens.d0=model.pp.cens.d0$data
check.cens.d0=simdata%>%left_join(dt.pp.cens.d0%>%mutate(flag=1,time=period)%>%select(id,time,flag))
sum(check.cens.d0$flag, na.rm = T)==dim(dt.pp.cens.d0)[1]
check.cens.d0%>%filter(id==6)
simdata%>%filter(id==6)

model.pp.cens.d1=readRDS("/Users/kevinying/TTEvignette/cense_model_d1.rds")
dt.pp.cens.d1=model.pp.cens.d1$data
check.cens.d1=simdata%>%left_join(dt.pp.cens.d1%>%mutate(flag=1,time=period)%>%select(id,time,flag))
sum(check.cens.d1$flag, na.rm = T)==dim(dt.pp.cens.d1)[1]
check.cens.d1%>%filter(id==6)
simdata%>%filter(id==6)


prep_PP_data$switch_models$switch_n0
model.pp.switch.n0=readRDS("/Users/kevinying/TTEvignette/weight_model_switch_n0.rds")
dt.pp.switch.n0=model.pp.switch.n0$data
check.switch.n0=simdata%>%left_join(dt.pp.switch.n0%>%mutate(flag=1,time=period)%>%select(id,time,switch,flag))
sum(check.switch.n0$flag, na.rm = T)==dim(dt.pp.switch.n0)[1]
check.switch.n0a=check.switch.n0%>%select(id,time,eligible, A, Y, C, switch,flag)
check.switch.n0a%>%filter(id==6)
simdata%>%filter(id==6)

prep_PP_data$switch_models$switch_n1
model.pp.switch.n1=readRDS("/Users/kevinying/TTEvignette/weight_model_switch_n1.rds")
dt.pp.switch.n1=model.pp.switch.n1$data
check.switch.n1=simdata%>%left_join(dt.pp.switch.n1%>%mutate(flag=1,time=period)%>%select(id,time,switch,flag))
sum(check.switch.n1$flag, na.rm = T)==dim(dt.pp.switch.n1)[1]
check.switch.n1a=check.switch.n1%>%select(id,time,eligible, A, Y, C, switch,flag)
check.switch.n1a%>%filter(id==6)
simdata%>%filter(id==6)
 

prep_PP_data$switch_models$switch_d0
model.pp.switch.d0=readRDS("/Users/kevinying/TTEvignette/weight_model_switch_d0.rds")
dt.pp.switch.d0=model.pp.switch.d0$data
check.switch.d0=simdata%>%left_join(dt.pp.switch.d0%>%mutate(flag=1,time=period)%>%select(id,time,switch,flag))
sum(check.switch.d0$flag, na.rm = T)==dim(dt.pp.switch.d0)[1]
check.switch.d0a=check.switch.d0%>%select(id,time,eligible, A, Y, C, switch,flag)
check.switch.d0a%>%filter(id==6)
simdata%>%filter(id==6)
 

prep_PP_data$switch_models$switch_d1
model.pp.switch.d1=readRDS("/Users/kevinying/TTEvignette/weight_model_switch_d1.rds")
dt.pp.switch.d1=model.pp.switch.d1$data
check.switch.d1=simdata%>%left_join(dt.pp.switch.d1%>%mutate(flag=1,time=period)%>%select(id,time,switch,flag))
sum(check.switch.d1$flag, na.rm = T)==dim(dt.pp.switch.d1)[1]
check.switch.d1a=check.switch.d1%>%select(id,time,eligible, A, Y, C, switch,flag)
check.switch.d1a%>%filter(id==6)
simdata%>%filter(id==6)

```
Conclusion: rows from the next period of the time of eligible till switching are included. The period that the subject is eligible (the first period of follow-up) is not include. If a subject is eligible at multiple time points, all rows from next period to each time of eligible till corresponding time of switch are included.  Each period is counted once even it can be included in multiple trials.

Now we run a weighted pooled logistic regression.
```{r}
dt.pp=NULL
for (i in 1:length(prep_PP_data$data)) {
  tmp <- read.csv(file.path(prep_PP_data$data[i]))
  dt.pp <- rbind(dt.pp, tmp)
}
pp_result=trial_msm(
  data=dt.pp,
  estimand_type='PP',
  outcome_cov = ~ X1 + X2 + X3 + X4 + age+followup_time,
  model_var = "assigned_treatment",
  #weight="weight",
  analysis_weights = "p99",
  weight_limits = c(0, Inf),
  include_followup_time = ~followup_time + I(followup_time^2),
  include_trial_period = ~trial_period + I(trial_period^2),
  glm_function="glm",
  quiet = TRUE
 )
 summary(pp_result, digits = 2)
```
