---
title: "A new-user, active comparator design; Intention-to-treat analysis"
author: "Kevin Ying, Peirong Hao, Yizhe Xu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: reference.bib  
link-citations: yes
objective: To apply the target trial emulation framework to create a analysis-ready data set
vignette: >
  %\VignetteIndexEntry{Target Trial Emulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(2009)
options(warn = -1)
```


# Introduction
##Peirong, add an introduction to A new-user, active comparator design; Intention-to-treat analysis

# Load packages
```{r, message = FALSE, warning = FALSE}
library(TrialEmulation)
library(dplyr)
library(tidyr)
library(ggplot2)
library(survival)
library(survminer)
library(lubridate)
source("R/generate observed data.R")
```
Simulate data
```{r}
# Set seed
set.seed(412)
#simulate data
simdata0=generate.obs.data(n=1000,J=20,beta0.trt0=1,beta.adherence=2, beta0.outcome=-5, beta.trt.effect=-1.2, beta0.censoring=-1)
simdata=simdata0%>%select(id, time, X1, X2, X3, X4, age, A, Y, C) 
```
Now we start to create a analysis-ready from the simulated data. We will emulate a randomized clinic trial described below. 
Add a table here to describe the trial design. 

First, we create an eligible indicator. Our inclusion criteria are: 1)age >=50, 2)no history of Y by current time, and 3)no treatment in previous 2 years of current time
```{r}
simdata <- simdata %>%
  group_by(id) %>%
  mutate(eligible = as.integer(age >= 50 & cumsum(Y) == 0 & (lag(A)+lag(A,2)) == 0)) %>%
  ungroup()
```
Let's take a look on a sample. 
```{r}
simdata%>%filter(id==2)

```
We can see that:  
At time 0 through 6, his age is below 50 and so e is not eligible for the study.
At time 7 through 9, his age is over 50 and he has no history of Y but he has treatment within previous 2 years, so he is not eligible.
At time 10 through 13, his age is 50 and he has no history of Y and no treatment in previous 2 years, so he is eligible.
At time 14 and after, he has treatment in previous 2 years, so he is not eligible.
Now we identify the subjects who are eligible ever and the first time they are eligible
```{r}
eligible_info <- simdata %>%
  group_by(id) %>%
  summarise(first_eligible_time = min(time[eligible == 1], na.rm = TRUE),
            ever_eligible = any(eligible == 1)) %>%
  ungroup()
```
How many ever eligible? 
```{r}
table(eligible_info$ever_eligible)
```
Check a few of them. 
```{r}
eligible_info %>% filter(id %in% 1:3)
```
As we explained above, subject 2 is first eligible at time 10. And we can also see that subject 1 and 3 are never eligible. 

Now merge the information
```{r}
simdata <- simdata %>%
  left_join(eligible_info, by = "id") 
```

now we are ready to do the analysis
We will perform intention-to-treatment (ITT) analysis and per-protocol (PP) analysis.
Explain ITT and PP here. 
#Case I:  A new-user, active comparator design; Intention-to-treat analysis
 
Subjects become eligible at multiple times. One option is just use the first time they are eligible. Below we prepare the data for analysis using this option.
In this option, we start follow up from the first eligible time
Select the eligible subjects and keep rows at and after first eligible time.
```{r}
simdata1 <- simdata %>%
  filter(time >= first_eligible_time) %>%
  select(-ever_eligible)
simdata.one.trial <- simdata1 %>%
  mutate(follow_up = time - first_eligible_time) %>%
  select(-first_eligible_time, -eligible) 
baseline=simdata.one.trial %>%filter(follow_up==0) %>%
  select(id, A, X1, X2, X3, X4, age) %>%
  rename(assigned_treatment = A, X1_0 = X1, X2_0 = X2, X3_0 = X3, X4_0 = X4, age_0 = age) 
simdata.one.trial=simdata.one.trial %>%
  left_join(baseline, by = "id") %>%
  arrange(id, follow_up)

```
This data is ready for ITT analysis.

To perform appropriate analysis we should take into consideration the censoring mechanism and the treatment assignment mechanism. Bellow are 4 different situations determined by the two mechanisms. 
1) treatment assignment can be seems random and censoring is non-informative.## Add example here.##. In this situation, a standard ITT analysis can be performed.
2) treatment assignment is not random and censoring is non-informative. ## Add example here.##. In this situation, we need to adjust the bias caused by confounding. We can use propensity score weighting or regression with confounding variable added as covariates to adjust the bias.
3) treatment assignment is random and censoring is informative. ## Add example here.##. In this situation, we can use inverse probability of censoring weighting (IPCW) to adjust the bias caused by informative censoring.
4) treatment assignment is not random and censoring is informative. ## Add example here.##. In this situation, we need to adjust the bias caused by confounding and informative censoring. We can use a combination of propensity score weighting of treatment and censoring to adjust the bias.

Below we demonstrate the analysis under each of the 4 situations.

#Situdation 1: treatment assignment can be seems random and censoring is non-informative.
We can do a Cox regression in this situation. 
We first convert the long format data (multiple rows per subject) to short format data (one row per subject). We keep the baseline values and the follow-up time and status
```{r}
#event time of subjects with event or censored
 event.cens.time= simdata.one.trial %>%filter(Y==1 | C==1) %>%
  group_by(id) %>%
  summarise(
    follow_up = min(follow_up, na.rm = TRUE),
 
  ) %>%
  ungroup()
event.cens.time=event.cens.time%>%left_join(simdata.one.trial)%>%
  select(id, assigned_treatment,X1_0, X2_0, X3_0, X4_0, age_0, follow_up, Y, C )%>%mutate(
    outcome = ifelse(Y == 1 & C==0, 1, 0) 
  ) 
#follow-up time of other subjects
event.cens.time0=simdata.one.trial %>%filter(!id%in%event.cens.time$id) %>%
  group_by(id) %>%
  summarise(
    follow_up = max(follow_up, na.rm = TRUE)
  ) %>%
  ungroup() %>%mutate(
    outcome = 0
  ) 
event.cens.time0=event.cens.time0%>%left_join(simdata.one.trial)%>%
  select(id, assigned_treatment,X1_0, X2_0, X3_0, X4_0, age_0, follow_up, outcome)
#combine the two data sets
simdata.short <- bind_rows(event.cens.time, event.cens.time0) %>%
  arrange(id)
#check if all subjects are included
setequal(unique(simdata.one.trial$id), unique(simdata.short$id))

```
Now we simply perform a Cox regression
```{r}
cox_model <- coxph(Surv(follow_up, outcome) ~ assigned_treatment, data = simdata.short)
summary(cox_model)
```
#situation 2: treatment assignment is not random and censoring is non-informative.
We can use propensity score weighting to adjust the bias caused by confounding. We can use the data we prepared above to do this analysis.
```{r}
# Fit a logistic regression model to estimate the propensity scores
ps_model <- glm(assigned_treatment ~ X1_0 + X2_0 + X3_0 + X4_0 + age_0, 
                 data = simdata.short, 
                 family = binomial)
# Calculate the propensity scores
p=predict(ps_model, type = "response")
# Calculate the inverse probability of treatment weights
simdata.short=simdata.short%>%mutate(weight=if_else(assigned_treatment==1, 1/p, 1/(1-p)))
#check overlapping of the weight
simdata.short%>%ggplot(aes(x=weight, fill=as.factor(assigned_treatment))) +
  geom_histogram(position="identity", alpha=0.5, bins=30) +
  labs(title="Distribution of Weights by Treatment Group", x="Weight", y="Count") +
  scale_fill_discrete(name="Treatment Group") +
  theme_minimal()
#truncate the weight at 97.5th percentile
simdata.short <- simdata.short %>%
  mutate(weight = ifelse(weight > quantile(weight, 0.975), quantile(weight, 0.975), weight))
###here to add code to check banalce of the covariates

#run a weighted cox regression
weighted_cox_model <- coxph(Surv(follow_up, outcome) ~ assigned_treatment, 
                             data = simdata.short, 
                             weights = weight)
summary(weighted_cox_model)
```
#situation 3: treatment assignment is random and censoring is informative.
We can use inverse probability of censoring weighting (IPCW) to adjust the bias caused by informative censoring. 

```{r}
# Fit a logistic regression model to estimate the censoring probabilities
censor_model <- glm(C ~ assigned_treatment + X1_0 + X2_0 + X3_0 + X4_0 + age_0, 
                     data = simdata.one.trial, 
                     family = binomial)
# Calculate the censoring probabilities
censor_probs <- predict(censor_model, type = "response")
# Calculate the inverse probability of censoring weights
simdata.one.trial <- simdata.one.trial %>%
  mutate(censor_weight = ifelse(C == 1, 1 / censor_probs, 1 / (1 - censor_probs)))
#check overlapping of the weight
simdata.one.trial %>%
  ggplot(aes(x = censor_weight, fill = as.factor(assigned_treatment))) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(title = "Distribution of Censoring Weights by Treatment Group", 
       x = "Censoring Weight", y = "Count") +
  scale_fill_discrete(name = "Treatment Group") +
  theme_minimal()
# Truncate the censoring weights at the 97.5th percentile
simdata.one.trial <- simdata.one.trial %>%
  mutate(censor_weight = ifelse(censor_weight > quantile(censor_weight, 0.975), 
                                quantile(censor_weight, 0.975), censor_weight))
###here to add code to check balance of the covariates
# Run a weighted pooled logistic regression
weighted_ipcw_model <- glm(Y ~ assigned_treatment+follow_up+follow_up^2, 
                             data = simdata.one.trial%>%filter(C!=0), #In an time interval, we consider censoring first. If censoring happens, then we don't kow the outcome. 
                             weights = censor_weight, 
                             family = binomial)
summary(weighted_ipcw_model)


```
situation 4: treatment assignment is not random and censoring is informative.
We can use a combination of propensity score weighting of treatment and censoring to adjust the bias.We can use the weight we calculated in situation 2 and 3 for this analysis
```{r}
simdata.one.trial=simdata.one.trial%>%left_join(simdata.short%>%select(id, weight)%>%rename(trt.weight=weight))%>%mutate(weight.TC=trt.weight*censor_weight)
weighted_iptw_ipcw_model <- glm(Y ~ assigned_treatment+follow_up+follow_up^2, 
                             data = simdata.one.trial%>%filter(C!=0), #In an time interval, we consider censoring first. If censoring happens, then we don't kow the outcome. 
                             weights = weight.TC, 
                             family = binomial)
summary(weighted_iptw_ipcw_model)