---
title: "Target Trial Emulation"
author: "Yizhe Xu"
date: "2025-04-01"
output: html_document
objective: To apply the target trial emulation framework to create a analysis-ready data set
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/u0918455/Library/CloudStorage/Box-Box/SDBC - Resources/TTE and Causal Inference Training/Week 11 - TTE coding")
```

# We use the exmaple data from the TrialEmulation R package 
```{r}
library(TrialEmulation)
library(dplyr)
library(lubridate)
```

## Case I: A new-user, active comparator design; Intention-to-treat analysis
Data structure: one-patient-per-row

1. Eligibility criteria -> This data set includes the study cohort of interest  
2. Treatment strategies -> The data set contains the treatment variable that clearly groups individuals
3. Treatment assignment -> The data set contains preselected baseline covariates (time-invariant) for controlling for confounding  
4. Follow-up            -> This data set clearly defines time zero (i.e., start of the follow-up) of the analysis and follow-up time 
5. Outcome              -> The data set contains time-to-event outcomes: an event indicator and the time to the event  

# Remove subjects who are not eligible 
# Eligible criteria: 1) No dementia at and prior to time zero, 2) index window is 2000-2017 
```{r}
data_censored_simp0 <- readRDS("./data_censored_simp4.rds")
View(data_censored_simp0)
# id --> patient identifier
# treatment --> indicator for receiving treatment in this period, 1=treatment, 0=non-treatment
# x3 --> A fixed categorical variable relating to treatment and the outcome
# x4 --> A fixed categorical variable relating to treatment and the outcome 

# Create a flag for eligibility
data_censored_simp1 <- data_censored_simp0 %>%
  mutate(eligible = ifelse(year(tx_init_date) <= 2017 & BL_dementia == 0, 1, 0))

# How many subjects are eligible?
table(data_censored_simp1$eligible)

# Remove ineligible subjects
data_censored_simp4 <- data_censored_simp1 %>% 
  filter(eligible == 1)
```

# Create the outcomes of interest: 
# 1) Dementia with death as the competing event
# 2) all-cause mortality
# 3) composite outcome of dementia and death

# There are 5 possibilities when we consider a non-terminal outcome with death as the competing event
#      dementia  death  loss-to-follow-up 
#  1)      0       0       1
#  2)      0       0       0   --> administratively censored 
#  3)      1       1       0
#  4)      1       0       0
#  5)      0       1       0
```{r}
# Create all-cause death outcome and (intermediate) dementia outcome
data_censored_simp5 <- data_censored_simp4 %>%
  mutate(ad_censor_days = as.numeric(as.Date("2022-12-31")-tx_init_date),
         
         death_days0 = ifelse(is.na(death_date)==FALSE, as.numeric(death_date-tx_init_date),  as.numeric(last_fu_date-tx_init_date)),
         death_days = pmin(death_days0, ad_censor_days),
         death = ifelse(death_days0 <= ad_censor_days & death==1, 1, 0),
         
         dementia_days0 = ifelse(is.na(dementia_date)==FALSE, as.numeric(dementia_date-tx_init_date), as.numeric(last_fu_date-tx_init_date)),
         dementia_days = pmin(dementia_days0, ad_censor_days),
         dementia = ifelse(dementia_days0 <= ad_censor_days & dementia==1, 1, 0))

# In real data, need to check if any dementia event occurs after the end of follow-up or death

# Create composite outcome and final dementia outcome with competing event of death
data_censored_simp6 <- data_censored_simp5 %>%
  mutate(death_dementia = ifelse(death==1|dementia==1, 1, 0),
         death_dementia_days = ifelse(death==1 & dementia==0, death_days, 
                                      ifelse(death==0 & dementia==1, dementia_days, pmin(dementia_days, death_days))),
         
         dementia_compete_death = ifelse(dementia==1, 1,
                                     ifelse(death==1, 2, 0)),
         dementia_compete_death_days = death_dementia_days)

```

# Save the clean data for downstream analyses 
```{r}
data_censored_simp_ITT <- data_censored_simp6 %>%
  dplyr::select(id, treatment, x3, x4, 
                death, death_days, death_dementia, death_dementia_days,
                dementia_compete_death, dementia_compete_death_days) 
saveRDS(data_censored_simp_ITT, file="./data_censored_simp_ITT.rds") 
```




## Case II: A new-user, active comparator design; Per-protocol analysis

Data structure: person-time long format 
```{r}
library(TrialEmulation)
data(data_censored) 

data_censored_simp_PP <- data_censored %>%
  mutate(period = period + 1,
         age = case_when(period==1 ~ age*12,
                         TRUE ~ age*12-11*(period-1))) %>%
  rename(month = period, event = outcome, LTFU = censored) %>%
  select(-c(age_s, eligible)) 

View(data_censored_simp_PP)
# id patient identifier
# month time period
# treatment indicator for receiving treatment in this period, 1=treatment, 0=non-treatment
# x1 A time-varying categorical variable relating to treatment and the outcome
# x2 A time-varying numeric variable relating to treatment and the outcome
# x3 A fixed categorical variable relating to treatment and the outcome
# x4 A fixed categorical variable relating to treatment and the outcome
# age patient age in months
# event event indicator 
# LTFU censoring indicator

# Issues in real data: 
# 1. Missing data in time-varying covariates
# 2. Missing entire visit months --> aggregate data: 3- or 6-month intervals 
```

# Create treatment adherence variable
# Consider a strict protocol of 1) adhering to initial treatment every month, 2) no contraindication is allowed
```{r}
tmp <- data_censored_simp_PP %>%
  group_by(id) %>%
  filter(row_number()==1) %>%
  rename(tx0 = treatment) %>%
  select(id, tx0)
  
data_censored_simp_PP2 <- data_censored_simp_PP %>% left_join(tmp, by = "id")
data_censored_simp_PP3 <- data_censored_simp_PP2 %>%
  mutate(adherence = ifelse(treatment == tx0, 1, 0))
View(data_censored_simp_PP3)
```

# Create artifically censored variable --> assume monotone censoring
# remove the rows of data after deviating from the protocol
```{r}
data_censored_simp_PP4 <- data_censored_simp_PP3 %>%
  mutate(art_censored = 1-adherence)
  
data_censored_simp_PP5  <- data_censored_simp_PP4 %>% 
  group_by(id) %>% 
  mutate(cum_art_censored = cumsum(art_censored)) %>% 
  filter(cum_art_censored %in% c(0,1)) %>% # We keep the row of being artificially censored for estimating the adherence probabilities 
  select(-cum_art_censored)

View(data_censored_simp_PP5)
```

## Case III: A new-user, placebo control design; Intention-to-treat analysis (TrialEmulation)

Data structure: person-trial long format 
```{r}
library(TrialEmulation)
data(data_censored)
View(data_censored)
# age_s standardized patient age

# Expand data using data_preparation 
# Assume random/independent censoring 
data_censored_ITT <- data_preparation(data_censored, estimand_type = "ITT", outcome_cov = ~x1+x2+x3+x4+age_s)
data_censored_ITT$min_period
data_censored_ITT$max_period # A sequence of 20 trials

# total number of subjects in the expanded data 
data_censored_ITT$N

# expanded data 
data_censored_ITT_dat <- data_censored_ITT$data
View(data_censored_ITT_dat)

data_censored_ITT_dat_id2 <- data_censored_ITT_dat %>% 
  filter(id == 2) %>%
  arrange(id, trial_period, followup_time)
View(data_censored_ITT_dat_id2)

data_censored_ITT_dat_id4 <- data_censored_ITT_dat %>% 
  filter(id == 4) %>%
  arrange(id, trial_period, followup_time)
View(data_censored_ITT_dat_id4)
```

# Explore dependent censoring
```{r}
data_censored_ITT2 <- data_preparation(data_censored, estimand_type = "ITT", outcome_cov = ~x1+x2+x3+x4+age_s,
                                       use_censor_weights = TRUE, cense = "censored", 
                                       cense_d_cov = ~ x1 + x2 + x3 + x4 + age_s, cense_n_cov = ~ x3 + x4,
                                       pool_cense = "numerator", # does not allow the numerator model to be fitted separately for the treatment groups
                                       glm_function = "parglm", nthreads = 4, method = "FAST", quiet = TRUE)

data_censored_ITT2$censor_models
```

# Case IV: A new-user, placebo control design; Per-protocol analysis

Data structure: person-trial-time long format 
```{r}
library(TrialEmulation)
data(data_censored)
View(data_censored)

# Expand data using data_preparation 
data_censored_PP <- data_preparation(data_censored, estimand_type = "PP", outcome_cov = ~x1+x2+x3+x4+age_s,
                                     switch_d_cov = ~ x1 + x2 + x3 + x4 + age_s + time_on_regime + I(time_on_regime^2),
                                     switch_n_cov = ~ x3 + x4 + time_on_regime + I(time_on_regime^2),
                                     use_censor_weights = TRUE, cense = "censored", 
                                     cense_d_cov = ~ x1 + x2 + x3 + x4 + age_s, cense_n_cov = ~ x3 + x4,
                                     pool_cense = "none", 
                                     glm_function = "parglm", nthreads = 4, method = "FAST", quiet = TRUE)
data_censored_PP$min_period
data_censored_PP$max_period # A sequence of 19 trials

# total number of subjects in the expanded data 
data_censored_PP$N  

# expanded data 
data_censored_PP_dat <- data_censored_PP$data

data_censored_PP_dat_id2 <- data_censored_PP_dat %>% 
  filter(id == 2)
View(data_censored_PP_dat_id2)

data_censored_PP_dat_id4 <- data_censored_PP_dat %>% 
  filter(id == 4) %>%
  arrange(id, trial_period, followup_time)
View(data_censored_PP_dat_id4)
```

# Explore the censoring and treatment models
```{r}
working_dir <- "/Users/u0918455/Library/CloudStorage/Box-Box/SDBC - Resources/TTE and Causal Inference Training/Week 11 - TTE coding"
data_censored_PP <- data_preparation(data_censored, estimand_type = "PP", outcome_cov = ~x1+x2+x3+x4+age_s,
                                     switch_d_cov = ~ x1 + x2 + x3 + x4 + age_s + time_on_regime + I(time_on_regime^2),
                                     switch_n_cov = ~ x3 + x4 + time_on_regime + I(time_on_regime^2),
                                     use_censor_weights = TRUE, cense = "censored", 
                                     cense_d_cov = ~ x1 + x2 + x3 + x4 + age_s, cense_n_cov = ~ x3 + x4,
                                     pool_cense = "none", 
                                     data_dir = working_dir, save_weight_models = TRUE,
                                     glm_function = "parglm", nthreads = 4, method = "FAST", quiet = TRUE)
# View(data_censored_PP$data)
# Note only baseline values of the time-varying variables are kept in the data set for downstream marginal structural models 
# The row of being artificially censored are not included 

# load the saved models 
weight_model_switch_d0 <- readRDS("weight_model_switch_d0.rds")
weight_model_switch_d1 <- readRDS("weight_model_switch_d1.rds")
weight_model_switch_n0 <- readRDS("weight_model_switch_n0.rds")
weight_model_switch_n1 <- readRDS("weight_model_switch_n1.rds")

cense_model_d0 <- readRDS("cense_model_d0.rds")
cense_model_d1 <- readRDS("cense_model_d1.rds")
cense_model_n0 <- readRDS("cense_model_n0.rds")
cense_model_n1 <- readRDS("cense_model_n1.rds")

# Prepare the data set for making predictions --> only looking at trial period zero for now 
tmp_dat <- data_censored_PP$data %>%
  group_by(id, trial_period) %>%
  mutate(adherence = ifelse(treatment == assigned_treatment, 1, 0), # in the cleaned PP data set, this adherence = 1 everywhere
         time_on_regime = cumsum(adherence)-1,  # -1 to not account the one at baseline 
         followup_time2 = followup_time+trial_period) %>%
  select(-c(x1, x2, age_s)) %>% 
  filter(trial_period == 0)

data_censored_tmp <- data_censored %>%
  rename(followup_time2 = period) %>%
  select(id, followup_time2, x1, x2, age_s)

tmp_dat2 <- tmp_dat %>% 
  left_join(data_censored_tmp, by=c("id", "followup_time2"))%>%
  arrange(id, trial_period, followup_time)


# for control arm
ctl_dat <- tmp_dat2 %>%
  filter(assigned_treatment == 0)

# probabilities of receiving treatment = 1 at each month 
weight_switch_d0 <- predict(weight_model_switch_d0, newdata = ctl_dat, type = "response")
weight_switch_n0 <- predict(weight_model_switch_n0, newdata = ctl_dat, type = "response")
switch_ctl <- (1-weight_switch_n0)/(1-weight_switch_d0) # 1-prob(tx = 1) = prob(tx = 0)

cense_d0 <- predict(cense_model_d0, newdata = ctl_dat, type = "response")
cense_n0 <- predict(cense_model_n0, newdata = ctl_dat, type = "response")
cense_ctl <- cense_n0/cense_d0

ctl_dat2 <- data.frame(ctl_dat, switch_ctl, cense_ctl) %>%
  group_by(id, trial_period) %>%
  mutate(prod_w = switch_ctl*cense_ctl,
         prod_w_new = case_when(followup_time == 0 ~ 1, # assuming no one is censored or non-adherent at baseline
                                TRUE ~ prod_w),
         cum_prod_w = cumprod(prod_w_new))

View(ctl_dat2[, c("id", "trial_period", "followup_time", "weight", "cum_prod_w")])


# for treatment arm 
trt_dat <- tmp_dat2 %>%
  filter(assigned_treatment == 1)

weight_switch_d1 <- predict(weight_model_switch_d1, newdata = trt_dat, type = "response")
weight_switch_n1 <- predict(weight_model_switch_n1, newdata = trt_dat, type = "response")
switch_trt <- weight_switch_n1/weight_switch_d1

cense_d1 <- predict(cense_model_d1, newdata = trt_dat, type = "response")
cense_n1 <- predict(cense_model_n1, newdata = trt_dat, type = "response")
cense_trt <- cense_n1/cense_d1

trt_dat2 <- data.frame(trt_dat, switch_trt, cense_trt) %>%
  group_by(id, trial_period) %>%
  mutate(prod_w = switch_trt*cense_trt,
         prod_w_new = case_when(followup_time == 0 ~ 1, 
                                TRUE ~ prod_w),
         cum_prod_w = cumprod(prod_w_new))

View(trt_dat2[, c("id", "trial_period", "followup_time", "weight", "cum_prod_w")])
```




# inconsistent component 
```{r}
ctl_dat3 <- data.frame(ctl_dat, switch_ctl, cense_ctl) %>%
  group_by(id, trial_period) %>%
  mutate(cense_ctl_new = case_when(followup_time == max(followup_time) ~ 1,
                                TRUE ~ cense_ctl),
         switch_ctl_new = case_when(followup_time == 0 ~ 1,
                                TRUE ~ switch_ctl),
         prod_w = switch_ctl_new*cense_ctl_new,
         cum_prod_w = cumprod(prod_w))

View(ctl_dat3[, c("id", "trial_period", "followup_time","cense_ctl", "cense_ctl_new", "switch_ctl", "switch_ctl_new", "weight", "cum_prod_w")])
```

